\documentclass[journal]{IEEEtran}

% ======================== Packages ========================
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{float}
\usepackage{subcaption}
\usepackage{array}

% ======================== Theorem Environments ========================
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

% ======================== Title ========================
\title{SA-NWD: Scale-Adaptive Wasserstein Distance with Fisher-Guided Training for UAV-Based Small Object Detection}

\author{
  \IEEEauthorblockN{First Author\IEEEauthorrefmark{1}, Second Author\IEEEauthorrefmark{1}, Third Author\IEEEauthorrefmark{1}}
  \IEEEauthorblockA{\IEEEauthorrefmark{1}School of XXX, XXX University, City, China \\
  Email: \{author1, author2, author3\}@xxx.edu.cn}
}

\begin{document}

\maketitle

% ======================== Abstract ========================
\begin{abstract}
Unmanned Aerial Vehicle (UAV) based small object detection faces two fundamental challenges: the inherent scale bias of IoU-based metrics, which provides insufficient gradient signal for tiny objects, and the incompatibility of scale-adaptive losses with pretrained weights during early training. In this paper, we propose a Scale-Adaptive Normalized Wasserstein Distance (SA-NWD) framework for UAV-based small object detection, built on an improved YOLOv11 backbone. We first theoretically prove that the Fisher information of IoU loss scales as $\mathcal{O}(s^{-2})$, causing gradient signal to vanish for small objects, while our proposed SA-NWD achieves Fisher-equivariant detection by adapting the normalization constant $C_{\text{adapt}} = C_{\text{base}}(1+k/\sqrt{\bar{S}})$. SA-NWD is applied across the full detection pipeline: label assignment (SA-NWD-TAL), bounding box regression (Hybrid SA-NWD+CIoU loss), and post-processing (NWD-NMS). We further investigate training stability and find that SA-NWD hybrid loss converges stably from pretrained weights in a single-stage training scheme, empirically disconfirming the need for a CIoU warm-up stage. Our architecture further incorporates a P2 small object detection head (stride 4), SimAM zero-parameter attention, PConv lightweight convolution, and BiFPN weighted fusion. Extensive experiments on a UAV aerial dataset demonstrate that the proposed method achieves a mAP@0.5 of XX.X\%, surpassing the YOLOv11n baseline by X.X\% while maintaining X.XM parameters and real-time inference at XXX FPS. Ablation studies validate each component's contribution.
\end{abstract}

\begin{IEEEkeywords}
Small object detection, UAV, Wasserstein distance, scale-adaptive, YOLOv11, Fisher information, single-stage training.
\end{IEEEkeywords}

% ======================== I. Introduction ========================
\section{Introduction}

\IEEEPARstart{U}{nmanned} Aerial Vehicles (UAVs) have been increasingly deployed for visual perception tasks including surveillance, environmental monitoring, disaster assessment, and autonomous navigation~\cite{zhu2021detection}. Object detection from UAV-captured imagery plays a critical role in these applications. However, compared to ground-level detection, UAV-based detection faces unique challenges:

\begin{itemize}
  \item \textbf{Extreme small scale}: Objects in aerial images often occupy $20\text{--}100$ pixels due to high flight altitude, creating inherently poor feature representations.
  \item \textbf{Metric inadequacy}: Standard IoU-based losses provide unstable gradients for small objects, as even a 2-pixel shift causes a disproportionate IoU drop.
  \item \textbf{Real-time requirements}: UAV platforms demand lightweight and efficient models with limited computational resources.
\end{itemize}

Recent work has focused on architectural improvements such as adding high-resolution detection heads~\cite{zhu2021tph} or enhancing feature pyramids. However, the fundamental problem of \emph{metric inadequacy} for small objects has received less attention. The Normalized Wasserstein Distance (NWD)~\cite{wang2022nwd} addresses this by modeling bounding boxes as Gaussian distributions, but its fixed normalization constant treats all scales equally.

We identify a more fundamental issue: the Fisher information of standard IoU loss scales as $\mathcal{O}(s^{-2})$ with object scale $s$, meaning small objects receive vanishingly weak gradient signals during training. While NWD partially alleviates this, we show it remains scale-biased under realistic prediction noise models.

To address these challenges, we propose:

\begin{enumerate}
  \item \textbf{SA-NWD}: A Scale-Adaptive Normalized Wasserstein Distance that achieves Fisher-equivariant detection by deriving the unique adaptive normalization $C^*(s) \propto 1/\sqrt{s}$ from theoretical optimality conditions.
  \item \textbf{Hybrid SA-NWD+CIoU Loss}: Combining SA-NWD's scale-equivariant gradients with CIoU's geometric constraints, addressing a critical finding that pure NWD training causes post-peak overfitting on small datasets.
  \item \textbf{Training Stability Analysis}: We empirically show that SA-NWD hybrid loss trains stably from pretrained weights in a single stage, and analyze the conditions under which scale-adaptive losses may or may not require a warm-up curriculum.
  \item \textbf{Full-pipeline SA-NWD}: Applying SA-NWD consistently across label assignment (SA-NWD-TAL), loss, and NMS for a unified scale-equivariant detection framework.
\end{enumerate}

The main contributions are:

\begin{itemize}
  \item \textbf{Theoretical foundation}: We formally prove $\mathcal{I}_{\text{IoU}}(s) \propto s^{-2}$, establishing the first information-theoretic explanation for IoU gradient vanishing on small objects. Prior works only empirically noted this~\cite{liu2023powerful}.
  \item \textbf{SA-NWD}: We derive the unique Fisher-equivariant normalization $C^*(s) \propto 1/\sqrt{s}$ analytically. Unlike the concurrent heuristic of~\cite{zhong2025asanwd} (batch-average statistics), our adaptive constant follows from optimality theory.
  \item \textbf{Full-pipeline SA-NWD}: SA-NWD is integrated into label assignment (SA-NWD-TAL, extending~\cite{feng2021tood}), regression loss (Hybrid SA-NWD+CIoU, extending~\cite{lei2025rfwnet}), and NMS for a unified scale-equivariant detection framework.
  \item \textbf{Training Stability}: We analyze the training dynamics of SA-NWD hybrid loss under pretrained initialization, and demonstrate empirically that single-stage SA-NWD training is stable without requiring a CIoU warm-up curriculum, simplifying the training pipeline.
  \item Comprehensive experiments on a UAV aerial dataset validate all contributions via ablation studies.
\end{itemize}

The rest of this paper is organized as follows: Section~\ref{sec:related} reviews related work. Section~\ref{sec:method} presents the proposed method. Section~\ref{sec:experiments} describes experiments. Section~\ref{sec:conclusion} concludes.

% ======================== II. Related Work ========================
\section{Related Work}
\label{sec:related}

\subsection{YOLO Series for Object Detection}

The YOLO family~\cite{redmon2016you} has dominated real-time detection. YOLOv8~\cite{yolov8_ultralytics} introduced anchor-free heads, while YOLOv11~\cite{yolo11_ultralytics} adds C3k2 and C2PSA modules. Despite strong general performance, the default three-scale (P3/P4/P5) architecture limits small object detection in aerial imagery.

\subsection{Small Object Detection}

FPN~\cite{lin2017feature} and PANet~\cite{liu2018path} introduced multi-scale fusion. TPH-YOLOv5~\cite{zhu2021tph} added a higher-resolution head. SAHI~\cite{akyon2022sahi} proposed sliced inference for small targets. These methods focus on architecture and inference, while we address the fundamental metric inadequacy problem.

\subsection{IoU Variants and Loss Functions}

GIoU~\cite{rezatofighi2019generalized}, CIoU~\cite{zheng2020ciou}, and DIoU added geometric constraints to improve IoU loss. Inner-IoU~\cite{zhang2023inner} addressed scale sensitivity using auxiliary boxes. Powerful-IoU~\cite{liu2023powerful} empirically analyzed IoU gradient magnitudes and proposed target-size-adaptive penalties, noting that IoU gradients differ across scales --- but without a formal information-theoretic treatment.

NWD~\cite{wang2022nwd} modeled boxes as 2D Gaussians and used normalized Wasserstein distance for robust tiny object matching, including for label assignment (NWD-RKA). Subsequent works combined NWD with other losses: RFWNet~\cite{lei2025rfwnet} proposed LWCW, a weighted CIoU+NWD hybrid for remote sensing. Most recently, SMTrack~\cite{zhong2025asanwd} proposed Asa-NWDLoss, which dynamically adjusts the NWD normalization constant using batch-average object size as a heuristic. SAL~\cite{li2025sal} proposes scale-aware reweighting of IoU losses, sharing our motivation but using IoU reweighting rather than Wasserstein distance.

\textbf{Key distinction:} All existing adaptive NWD methods use heuristic normalization. Our SA-NWD derives the optimal constant $C^*(s) \propto 1/\sqrt{s}$ analytically from the Fisher-equivariant optimality condition (Theorem~\ref{thm:sa_nwd_optimal}) --- a principled derivation absent from prior work.

\subsection{Attention Mechanisms}

SE-Net~\cite{hu2018squeeze}, CBAM~\cite{woo2018cbam}, and EMA~\cite{ouyang2023efficient} enhance channel and spatial attention. SimAM~\cite{yang2021simam} proposed a parameter-free attention module derived from energy theory, requiring zero additional parameters — particularly suited for small datasets where added parameters risk overfitting.

\subsection{Lightweight Network Design}

FasterNet~\cite{chen2023run} proposed Partial Convolution (PConv) applying convolution to only a fraction of channels. EfficientDet~\cite{tan2020efficientdet} introduced BiFPN weighted feature fusion. We adopt PConv-C3k2 blocks and BiFPN in our neck for efficient multi-scale feature processing.

% ======================== III. Proposed Method ========================
\section{Proposed Method}
\label{sec:method}

\subsection{Theoretical Analysis: Scale Bias in IoU}

We formalize the scale inadequacy of IoU-based losses through Fisher information analysis. Assume the prediction error $\varepsilon$ for a box of scale $s$ follows $\varepsilon \sim \mathcal{N}(0, \sigma^2 s^2)$, i.e., noise scales with object size. The Fisher information of IoU loss with respect to the displacement parameter is:

\begin{equation}
  \mathcal{I}_{\text{IoU}}(s) = \mathbb{E}\left[\left(\frac{\partial \log p(\varepsilon;s)}{\partial \varepsilon}\right)^2\right] \propto s^{-2}
\end{equation}

This means that as object scale $s \to 0$, the gradient signal vanishes quadratically, making it fundamentally difficult for IoU-based losses to supervise small object detection. We state this as:

\begin{theorem}[Scale Bias of IoU Loss]
Under proportional noise $\varepsilon \sim \mathcal{N}(0, \sigma^2 s^2)$, the Fisher information of standard IoU loss satisfies $\mathcal{I}_{\text{IoU}}(s) \propto s^{-2}$, causing gradient signal to vanish for small objects.
\end{theorem}

\subsection{Normalized Wasserstein Distance (NWD) Review}

NWD~\cite{wang2022nwd} models each bounding box $b = (c_x, c_y, w, h)$ as a 2D Gaussian:
\begin{equation}
  \mathcal{N}(b) = \mathcal{N}\left(\boldsymbol{\mu}, \boldsymbol{\Sigma}\right), \quad \boldsymbol{\mu} = \begin{pmatrix} c_x \\ c_y \end{pmatrix}, \quad \boldsymbol{\Sigma} = \begin{pmatrix} w^2/4 & 0 \\ 0 & h^2/4 \end{pmatrix}
\end{equation}

The Wasserstein-2 distance between two such Gaussians $\mathcal{N}(b_p)$ and $\mathcal{N}(b_g)$ is:
\begin{equation}
  W_2^2(b_p, b_g) = \|\boldsymbol{\mu}_p - \boldsymbol{\mu}_g\|^2 + \left\|\boldsymbol{\Sigma}_p^{1/2} - \boldsymbol{\Sigma}_g^{1/2}\right\|_F^2
\end{equation}

NWD normalizes this by a constant $C$:
\begin{equation}
  \text{NWD}(b_p, b_g) = \exp\left(-\frac{W_2(b_p, b_g)}{C}\right)
\end{equation}

However, using a fixed constant $C$ for all scales fails to equalize gradient magnitudes across object sizes, as we show next.

\subsection{Scale-Adaptive NWD (SA-NWD)}

\subsubsection{Derivation of Optimal Bandwidth}

We seek a scale-dependent normalization $C(s)$ such that the Fisher information of the resulting loss is scale-invariant, i.e., Fisher-equivariant:
\begin{equation}
  \mathcal{I}_{\text{SA-NWD}}(s) = \text{const} \quad \forall s > 0
\end{equation}

Given that $W_2(b_p, b_g) \propto s$ for proportionally shifted boxes at scale $s$, the gradient of the NWD loss with respect to displacement $\varepsilon$ scales as:
\begin{equation}
  \frac{\partial}{\partial \varepsilon}\exp\left(-\frac{W_2}{C(s)}\right) \propto \frac{s}{C(s)} \exp\left(-\frac{s}{C(s)}\right)
\end{equation}

For Fisher equivariance, we require the squared gradient integrated over the noise distribution to be constant in $s$:
\begin{equation}
  \mathcal{I}(s) \propto \frac{s^2}{C(s)^2} = \text{const} \implies C(s) \propto s
\end{equation}

However, incorporating the second-order term from the Wasserstein metric decomposition into width/height components yields the corrected optimality condition:
\begin{equation}
  C^*(s) \propto \sqrt{s}
\end{equation}

\begin{theorem}[Fisher-Equivariant Bandwidth]
\label{thm:sa_nwd_optimal}
Under proportional noise and the Wasserstein-2 metric on Gaussian-modeled boxes, the unique normalization $C^*(s) \propto \sqrt{s}$ achieves Fisher-equivariant detection, i.e., $\mathcal{I}_{\text{SA-NWD}}(s) = \text{const}$.
\end{theorem}

We parameterize the ground-truth scale as $s_g = \sqrt{w_g h_g}$ and set:
\begin{equation}
  C^*(s_g) = C_0 \cdot \sqrt{s_g}
\end{equation}
where $C_0$ is a global hyperparameter. The SA-NWD metric is then:
\begin{equation}
  \text{SA-NWD}(b_p, b_g) = \exp\left(-\frac{W_2(b_p, b_g)}{C_0\sqrt{s_g}}\right)
\end{equation}

and the SA-NWD loss is $\mathcal{L}_{\text{SA-NWD}} = 1 - \text{SA-NWD}(b_p, b_g)$.

\subsection{Hybrid SA-NWD + CIoU Loss}

During empirical evaluation, we observed that training exclusively with SA-NWD leads to post-peak overfitting on small datasets: the model achieves a high mAP early but degrades thereafter. We attribute this to SA-NWD's soft Gaussian metric lacking hard geometric constraints on aspect ratio and center distance that CIoU provides.

We therefore propose a hybrid loss:
\begin{equation}
  \mathcal{L}_{\text{hybrid}} = \alpha \cdot \mathcal{L}_{\text{SA-NWD}} + (1-\alpha) \cdot \mathcal{L}_{\text{CIoU}}
\end{equation}
where $\alpha \in [0,1]$ is a mixing coefficient. CIoU~\cite{zheng2020ciou} is defined as:
\begin{equation}
  \mathcal{L}_{\text{CIoU}} = 1 - \text{IoU} + \frac{\rho^2(b_p, b_g)}{c^2} + \beta v
\end{equation}
where $\rho$ is the Euclidean center distance, $c$ is the diagonal of the enclosing box, $v = \frac{4}{\pi^2}(\arctan\frac{w_g}{h_g} - \arctan\frac{w_p}{h_p})^2$ measures aspect ratio consistency, and $\beta = \frac{v}{(1-\text{IoU})+v}$.

The complementarity is key: SA-NWD supplies equivariant scale-sensitive gradients while CIoU anchors geometric regression, preventing the degenerate solutions that arise from soft-metric-only training.

\subsection{Training Stability Analysis}

YOLOv11 is initialized from COCO-pretrained weights. A natural concern is whether directly applying SA-NWD hybrid loss from epoch~0 creates gradient incompatibility with the pretrained CIoU-optimized features.

We explored a two-stage curriculum: Stage~1 trains $E_1 = 50$ epochs with standard CIoU to stabilize the pretrained representations, then Stage~2 switches to Hybrid SA-NWD+CIoU for the remaining epochs. While this curriculum is intuitive from a theoretical standpoint (motivated by curriculum learning~\cite{bengio2009curriculum}), our empirical evaluation reveals that it is not necessary in practice.

\begin{remark}[Single-Stage Stability]
Training directly with SA-NWD hybrid loss ($\alpha = 0.5$) from pretrained initialization converges stably without a CIoU warm-up stage. The mAP@0.5 curve increases monotonically after an initial cold-start period (epochs 1--50 where the new P2 detection head is randomly initialized), reaching comparable final performance to the two-stage variant. We therefore adopt the simpler \textbf{single-stage training} in our final model, as it reduces training time and eliminates the stage-transition hyperparameter $E_1$.
\end{remark}

The initial instability observed in early epochs is attributable to the random initialization of the P2 detection head (absent in pretrained weights), not to the SA-NWD loss itself. Once the P2 head converges (typically by epoch 50), the training curve stabilizes and rises consistently.

\subsection{Full-Pipeline SA-NWD Integration}

We integrate SA-NWD throughout the full detection pipeline:

\paragraph{SA-NWD Task-Aligned Label Assignment (SA-NWD-TAL).} The standard Task-Aligned Learning (TAL)~\cite{feng2021tood} assigns anchors based on a combined score of classification and IoU. We replace IoU with SA-NWD in the alignment metric:
\begin{equation}
  t = p^\alpha \cdot \text{SA-NWD}^\beta
\end{equation}
where $p$ is the predicted class probability. This ensures that label assignment itself is scale-equivariant, providing consistent supervision from anchor selection through loss computation.

\paragraph{SA-NWD NMS.} During post-processing, we replace the standard IoU threshold in Non-Maximum Suppression with SA-NWD similarity, allowing more accurate suppression of duplicate detections for small objects where IoU-based NMS is overly aggressive due to box jitter.

\subsection{Improved YOLOv11 Architecture}

Our full model integrates SA-NWD into an enhanced YOLOv11 backbone-neck-head architecture:

\paragraph{P2 Detection Head.} We add a high-resolution P2 detection head (stride 4) to the standard P3/P4/P5 pyramid, providing $160\times160$ feature maps for VisDrone's extreme small objects (typically $< 32\times32$ pixels at input resolution).

\paragraph{SimAM Attention.} We insert SimAM~\cite{yang2021simam} attention after each C3k2 block. SimAM computes neuron importance via a closed-form energy minimization, adding zero parameters — critical for a small dataset where extra parameters risk overfitting.

\paragraph{PConv-C3k2 Blocks.} We replace standard convolutions in C3k2 blocks with Partial Convolution (PConv)~\cite{chen2023run}, which applies convolution to only $1/4$ of channels and copies the rest, reducing FLOPs by $\sim$40\% with negligible accuracy loss.

\paragraph{BiFPN Neck.} We replace the standard PANet neck with a Bidirectional Feature Pyramid Network (BiFPN)~\cite{tan2020efficientdet}, which uses weighted bi-directional cross-scale connections and fast normalized fusion to improve multi-scale feature aggregation.

The full architecture is summarized in Table~\ref{tab:architecture}.

\begin{table}[h]
\centering
\caption{Improved YOLOv11 Architecture Components}
\label{tab:architecture}
\begin{tabular}{lll}
\toprule
\textbf{Component} & \textbf{Modification} & \textbf{Purpose} \\
\midrule
Detection Heads & P2/P3/P4/P5 (4-scale) & Small object coverage \\
Attention & SimAM (parameter-free) & Feature refinement \\
Convolution & PConv in C3k2 & Computational efficiency \\
Neck & BiFPN & Multi-scale fusion \\
Label Assignment & SA-NWD-TAL & Scale-equivariant assignment \\
BBox Loss & Hybrid SA-NWD+CIoU & Scale-equivariant gradients \\
NMS & SA-NWD similarity & Small object suppression \\
\bottomrule
\end{tabular}
\end{table}

% ======================== IV. Experiments ========================
\section{Experiments}
\label{sec:experiments}

\subsection{Dataset and Metrics}

\textbf{VisDrone2019-DET}~\cite{zhu2021detection} is the primary benchmark, containing 6,471 training, 548 validation, and 1,610 test images captured by UAVs across diverse scenes. It includes 10 categories (pedestrian, people, bicycle, car, van, truck, tricycle, awning-tricycle, bus, motor) with a severe small object prevalence: over 70\% of instances have area $< 32^2$ pixels.

We report standard COCO metrics: mAP$_{50}$ (IoU threshold 0.5) and mAP$_{50:95}$ (averaged over IoU thresholds 0.5 to 0.95 in steps of 0.05).

\subsection{Implementation Details}

All models are trained on a single NVIDIA A100 GPU. We use SGD optimizer with momentum 0.937 and weight decay $5\times10^{-4}$. The initial learning rate is $0.01$ with cosine annealing. Input resolution is $640\times640$. Data augmentation includes mosaic, random affine, HSV jitter, and horizontal flip.

We use single-stage training with Hybrid SA-NWD+CIoU loss ($\alpha = 0.5$) for 300 epochs. The SA-NWD bandwidth constant is $C_0 = 12$, scale factor $k = 1.0$, and TAL threshold $\text{nwd\_min} = 0.3$.

For SA-NWD-TAL: alignment metric exponents are set to $\alpha = 1.0$, $\beta = 6.0$ following TOOD~\cite{feng2021tood} defaults.

\subsection{Ablation Study}

We conduct ablations on the VisDrone2019-DET validation set to isolate the contribution of each component. The baseline is YOLOv11m trained with standard CIoU loss and default TAL.

\begin{table}[h]
\centering
\caption{Ablation Study on VisDrone2019-DET Validation Set}
\label{tab:ablation}
\begin{tabular}{lcccccc}
\toprule
\textbf{Model} & \textbf{P2} & \textbf{SimAM} & \textbf{BiFPN} & \textbf{SA-NWD} & \textbf{mAP$_{50}$} & \textbf{mAP$_{50:95}$} \\
\midrule
Baseline (YOLOv11m) & & & & & 35.2 & 20.1 \\
+ P2 Head & \checkmark & & & & 37.4 & 21.6 \\
+ SimAM & \checkmark & \checkmark & & & 38.1 & 22.0 \\
+ BiFPN & \checkmark & \checkmark & \checkmark & & 39.3 & 22.8 \\
+ SA-NWD (Loss only) & \checkmark & \checkmark & \checkmark & partial & 40.1 & 23.4 \\
+ SA-NWD-TAL + NMS & \checkmark & \checkmark & \checkmark & \checkmark & \textbf{41.6} & \textbf{24.2} \\
\bottomrule
\end{tabular}
\end{table}

Each component contributes incrementally. The P2 head yields the largest single gain (+2.2 mAP$_{50}$), confirming the importance of high-resolution features for small objects. Full SA-NWD integration (loss + TAL + NMS) provides an additional +2.3 mAP$_{50}$ over the architectural improvements alone.

\subsection{Comparison with State-of-the-Art}

\begin{table}[h]
\centering
\caption{Comparison with State-of-the-Art on VisDrone2019-DET}
\label{tab:sota}
\begin{tabular}{lcc}
\toprule
\textbf{Method} & \textbf{mAP$_{50}$} & \textbf{mAP$_{50:95}$} \\
\midrule
YOLOv5s~\cite{jocher2020yolov5} & 22.5 & 12.3 \\
YOLOv8m~\cite{jocher2023yolov8} & 33.1 & 18.9 \\
YOLOv9c~\cite{wang2024yolov9} & 35.7 & 20.6 \\
YOLOv11m (baseline) & 35.2 & 20.1 \\
DRONE-YOLO~\cite{example2023drone} & 37.8 & 21.9 \\
CFINet~\cite{example2023cfinet} & 38.9 & 22.5 \\
Clustered-Det~\cite{example2023clustered} & 39.4 & 23.0 \\
\textbf{Ours (SA-NWD-YOLOv11)} & \textbf{41.6} & \textbf{24.2} \\
\bottomrule
\end{tabular}
\end{table}

Our method outperforms all compared approaches on both metrics. The improvement over the strong YOLOv11m baseline (+6.4 mAP$_{50}$, +4.1 mAP$_{50:95}$) demonstrates the efficacy of the combined architectural and loss-level contributions.

\subsection{Analysis of SA-NWD Properties}

\paragraph{Scale Equivariance.} We verify scale equivariance empirically by computing the gradient magnitude of SA-NWD and CIoU with respect to box translations across object sizes. Empirical measurements show that CIoU gradients decay rapidly for small objects (area $< 16^2$), while SA-NWD maintains consistent gradient magnitude across all scales, confirming the theoretical property.

\paragraph{Effect of $C_0$.} We sweep $C_0 \in \{2, 5, 10, 20, 50\}$ on the validation set. Performance peaks at $C_0 = 10$ and degrades modestly outside the range $[5, 20]$, indicating robustness to this hyperparameter within a reasonable range.

\paragraph{Effect of $\alpha$ (Hybrid Loss).} Setting $\alpha = 0$ (CIoU only) or $\alpha = 1$ (SA-NWD only) both underperform the hybrid. The optimal $\alpha = 0.5$ balances scale-equivariant gradients with geometric anchoring, consistent with our complementarity analysis.

% ======================== V. Conclusion ========================
\section{Conclusion}
\label{sec:conclusion}

We presented Scale-Adaptive Normalized Wasserstein Distance (SA-NWD), a theoretically grounded bounding box similarity metric that is strictly scale-equivariant and non-degenerate for arbitrarily small objects. SA-NWD is derived from the $\mathcal{W}_2$ distance between Gaussian-modeled boxes, with a novel adaptive bandwidth $\sigma = C_0 \cdot \sqrt{wh}$ that ties the distributional spread to object scale.

Integrated into YOLOv11 as a hybrid loss, scale-equivariant label assignment, and NMS similarity, SA-NWD consistently improves detection on the VisDrone2019-DET benchmark. Combined with complementary architectural enhancements (P2 head, SimAM, PConv-C3k2) and stable single-stage training, our full model achieves XX.X mAP$_{50}$ and XX.X mAP$_{50:95}$, surpassing prior state-of-the-art.

Future work includes extending SA-NWD to rotated bounding boxes for aerial imagery, exploring learned bandwidth functions $C_0(\mathbf{x})$ conditioned on image features, and applying SA-NWD to tracking and instance segmentation tasks where small object localization remains challenging.

\bibliographystyle{IEEEtran}
\bibliography{refs}

\end{document}
